---
title: "Analyzing US Births and Basketball Recruits"
author: "Sakith"
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(readxl)
```

```{r}
#| message: false
#| cache: true
us_births <- read_excel("data/us_births_1994_2014.xlsx")
births_tibble <- us_births |>
  mutate(day_of_week = factor(day_of_week,
                              levels = c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat"),
                              ordered = TRUE))
summary(births_tibble$day_of_week)

births_tibble |>
  group_by(month, date_of_month) |>
  summarize(avg_births = mean(births), .groups = "drop") |>
  ggplot(aes(x = date_of_month, y = month, fill = avg_births)) +
  geom_tile() +
  scale_y_discrete(limits = month.name, name = "Month") +
  scale_x_continuous(breaks = 1:31, name = "Date of month") +
  labs(fill = "Mean births") +
  theme_minimal()

christmas_data <- births_tibble |>
  filter(month == 12, date_of_month %in% c(20:24, 25, 27:30)) |>
  group_by(year) |>
  summarize(
    dec_25_births = sum(births[date_of_month == 25]),
    baseline_births = mean(births[date_of_month != 25]),
    percent_of_baseline = dec_25_births / baseline_births * 100,
    .groups = "drop"
  )

christmas_day_of_week <- births_tibble |>
  filter(month == 12, date_of_month == 25) |>
  select(year, day_of_week)

christmas_data <- christmas_data |>
  left_join(christmas_day_of_week, by = "year")

summary(christmas_data)

christmas_data |>
  ggplot(aes(x = year, y = percent_of_baseline, color = day_of_week)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "December 25 births as % of baseline", color = "Day of week") +
  theme_minimal()

births_model <- lm(births ~ factor(year) + factor(month) + day_of_week, data = births_tibble)
summary(births_model)$r.squared

births_adjusted <- births_tibble |>
  mutate(pct_resid = residuals(births_model) / mean(births) * 100)

summary(births_adjusted$pct_resid)

calendar_resid <- births_adjusted |>
  filter(!(month == 2 & date_of_month == 29)) |>
  group_by(month, date_of_month) |>
  summarize(mean_pct_resid = mean(pct_resid), .groups = "drop") |>
  mutate(calendar_date = as.Date(paste("2001", month, date_of_month, sep = "-"), format = "%Y-%m-%d"))

summary(calendar_resid)

holiday_dips <- tibble(
  holiday = c("New Year's Day", "Independence Day", "Thanksgiving", "Christmas"),
  calendar_date = as.Date(c("2001-01-01", "2001-07-04", "2001-11-22", "2001-12-25"))
) |>
  left_join(calendar_resid, by = "calendar_date")

calendar_resid |>
  ggplot(aes(x = calendar_date, y = mean_pct_resid)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(data = holiday_dips, aes(x = calendar_date, y = mean_pct_resid), color = "red") +
  geom_text(
    data = holiday_dips,
    aes(x = calendar_date, y = mean_pct_resid, label = holiday),
    vjust = -0.6,
    hjust = 0.5,
    check_overlap = TRUE
  ) +
  labs(x = "Calendar date", y = "Mean percent residual") +
  theme_minimal()

```{r}
#| cache: true
basketball_tibble <- read_excel("data/nba_recruits.xlsx") |>
  mutate(
    tier = factor(tier,
                  levels = c("Never played", "Brief career", "Solid career", "All-Star level", "Superstar"),
                  ordered = TRUE),
    recruit_group = factor(recruit_group,
                           levels = c("#1–10", "#11–25", "#26–50", "#51–100", "Outside top 100"),
                           ordered = TRUE)
  )

glimpse(basketball_tibble)

basketball_tibble |>
  count(tier, name = "player_count") |>
  arrange(desc(player_count))

basketball_tibble |>
  select(rank, nba_mean_ws48, top_mean_wa, total_seasons, drafted) |>
  summary()
```

```{r}
basketball_tibble |>
  filter(recruit_group %in% c("#1–10", "#11–25", "#26–50", "#51–100")) |>
  count(recruit_group, tier, name = "n_players") |>
  ggplot(aes(y = recruit_group, x = n_players, fill = tier)) +
  geom_col(position = "fill") +
  labs(
    x = "Proportion of players",
    y = "Recruit group",
    fill = "Career tier"
  ) +
  theme_minimal()
```

```{r}
basketball_tibble |>
  filter(!is.na(nba_mean_ws48)) |>
  ggplot(aes(y = recruit_group, x = nba_mean_ws48, fill = recruit_group)) +
  geom_boxplot() +
  labs(
    x = "Career Win Shares per 48 minutes",
    y = "Recruit group",
    fill = "Recruit group"
  ) +
  theme_minimal()
```

```{r}
basketball_tibble |>
  filter(!is.na(nba_mean_ws48)) |>
  arrange(desc(nba_mean_ws48)) |>
  select(name, rank, recruit_group, nba_mean_ws48) |>
  slice_head(n = 10)
```

```{r}
label_data <- basketball_tibble |>
  filter(name %in% c("LeBron James", "Draymond Green", "Klay Thompson", "Gilbert Arenas"))

basketball_tibble |>
  filter(!is.na(rank), !is.na(top_mean_wa)) |>
  ggplot(aes(x = rank, y = top_mean_wa, color = tier)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
  geom_text(
    data = label_data,
    aes(label = name),
    vjust = -0.5,
    check_overlap = TRUE
  ) +
  labs(
    x = "Recruit rank",
    y = "Top career Wins Added",
    color = "Career tier"
  ) +
  theme_minimal()
```
```
